---
title: "New Zealand"
---

```{r}
#| label: plotting-ggplot2-nz-libs-load
#| include: false
library(colorspace)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(magrittr)
library(sf)
library(tibble)
library(mapswrData)
here::i_am("plotting/033-plotting.qmd")
```

# Objectives

The objective in this section is to produce a choropleth map of New Zealand with `ggplot2`.

# Introduction

Up until now, the `ggplot2` examples were demonstrating basic principles in map building.  Here, we'll build our first real choropleth map using New Zealand as an example.  We'll go through the full production pipline by (1) importing an `sf` data frame, (2) cutting the variable into bins, (3) labeling the factor variable and (4) plotting multiple layers of the map.

# Packages Required

```{r}
#| label: plotting-ggplot2-text-pkgs-libs-show
#| include: true
#| eval: false
library(colorspace)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(magrittr)
library(sf)
library(tibble)
library(mapswrData)
```

# Create

The first step is to load the `nz_business` data from the `mapswrData` package.

```{r}
data(nz_business)
nz_business %>% st_drop_geometry() %>% slice_head(n = 5)
```


# Business Indicators

The data comes from the official New Zealand Statistics agency.  The metadata state that it is "a monthly count of enterprises as at the end of each month, mostly sourced from administrative data" and includes only "economically significant enterprises. It can be an individual, private-sector and public-sector enterprises." These enterprises have a "GST turnover greater than $30,000 per year" [@statsnz_enterprises_2024]. The column "pct_business" will be assigned to a bin for the choropleth map.

```{r}
#| label: plotting-ggplot-nz-bi
nz_business %>% 
mutate(pct_bus_f = cut_interval(pct_business, n = 4, dig.lab = 1), 
       .before = geometry) -> nz_bus
levels(nz_bus$pct_bus_f) <- c(".61% to 9.1%",
                               "9.1% to 18%",
                               "18% to 26%",
                               "26% to 35%")
```

# Choropleth

```{r}
#| label: fig-plotting-ggplot-nz-bi
#| fig-cap: "May 2024 New Zealand Percentage of Businesses by District. Auckland, the home of the largest city, leads with 34.7% of the nation's businesses. Source is the New Zealand National Statistics. (If you're from New Zealand, this may not be all that enlightening!)"

values = colorspace::sequential_hcl(4, palette = "TealGrn", rev = T) #<1>
ggplot() +
geom_sf(data = nz_bus, mapping = aes(fill = pct_bus_f), 
        lwd = .5, col = "gray70") +
scale_fill_manual(name = "Pct.Total\nBusinesses", values = values) +
geom_label_repel(data = nz_bus, #<2>
                mapping = aes(x = lng, y = lat, label = region),
                force = 10,
                fill = alpha(c("white"),0.75),
                label.padding = .2,
                box.padding = .6,
                min.segment.length = 1.2,
                arrow = arrow(length = unit(0.01, "npc"),
                              type = "open"),
                size = 4) + #<2>
theme_void() +
theme( #<3>
    legend.text = element_text(size = 11),
    legend.title = element_text(margin = margin(t = 0, r = 0, b = 4, l = 0, unit = "pt"), size = 12),
    legend.box.margin = margin(t = 0, r = 5, b = 0, l = 5, unit = "pt"),
    legend.key.spacing.x = unit(1, units = "pt"),
    legend.key.spacing.y = unit(2, units = "pt"),
    
    plot.background = element_rect(
        fill = "white",
        colour = "black",
        linewidth = .5
  ) #<3>
)

```
1. Use `rev = T` to have dark color indicate high intensity.
2. Requires a lot of time for attractive labels.
3. Most changes dealt with legend appearance.

# Key Points

- Geometries will often include unexpected territories and features that require design choices as to whether they should be included or omitted

- Factor variables may need to be relabeled for attractive legends

- The `colorspace` package includes a `rev = ` argument that can be helpful in reversing the color palette

- Attractive annotations in the form of labels and text are time intensive

- Alter theme defaults only at the end of the project

<hr>
:::{#refs-plotting-024-rspatial}
:::
