---
title: "tmap"
---


```{r}
#| label: plotting-tmap-intro-libs-load
#| include: false
library(rnaturalearth)
library(dplyr)
library(magrittr)
library(sf)
library(tmap)
library(tmaptools)
library(withr)
here::i_am("plotting/040-plotting.qmd")
data(World)
```

# Objectives

The objective for this section is to introduce the `tmap` functions for plotting spatial data.

# Introduction

According to the `tmap` website, it is used for drawing thematic maps and is based on "A Layered Grammar of Graphics" like `ggplot2`. [@tennekes_tmap_2018]. "A grammar provides a strong foundation for understanding a diverse range of graphics. A grammar may also help guide us on what a well-formed or correct graphic looks like . . . " [@wickham_layered_2010] Layers are added using the `+` operator just like in `ggplot2`. Most functions begin with the `tm_` prefix. A great introduction is the package's getting started vignette. The vignette can be seen with the following code: `vignette("tmap-getstarted", package = "tmap")`. With `tmap` you can view the map in either static or interactive mode.  We'll take up interactive plots later and focus on static maps. `tmap` is the primary plotting package in the "Geocomputation with R" book with the rationales being:

>tmap is a powerful and flexible map-making package with sensible defaults. It has a concise syntax that allows for the creation of attractive maps with minimal code which will be familiar to ggplot2 users. It also has the unique capability to generate static and interactive maps using the same code via tmap_mode(). Finally, it accepts a wider range of spatial classes (including sf and terra objects) than alternatives such as ggplot2. [@lovelace2019geocomputation]

We’ll begin with a simple square and then move to an example of multiple squares.

:::{.callout-warning}
# Development Version

These lessons were built with the development version of `tmap` which is moving from version 3 to 4.  See the [GitHub repo](https://github.com/r-tmap/tmap) for latest development version and installation instructions.

:::

# Packages Required

```{r}
#| label: plotting-tmap-intro-libs-show
#| include: true
#| eval: false
library(rnaturalearth)
library(dplyr)
library(magrittr)
library(sf)
library(tmap)
library(tmaptools)
```

# First Plot - Square

## Create Object

To get started, a square polygon will be created at the (0, 0), the intersection of the Prime Meridian and the Equator.  Each side will be one unit in length and because the CRS "WGS 84" will be assigned, that unit is a degree.

```{r}
#| label: plotting-tmap-p1-sq-create
sf::st_polygon(list(rbind(c(0,0), c(0,1), c(1,1), c(1,0), c(0,0)))) %>%
st_sfc(., crs = 4326) %>%
st_sf(name = "square", .) -> square
```


## Mode

`tmap` has two modes: "plot" and "view".  The "plot" mode is for static plots and the "view" mode is for interactive plots.  The default is "plot".

```{r}
#| eval: false
tmap_mode("plot") #static
tmap_mode("view") #interactive
```


## Plot

```{r}
#| label: fig-plotting-tmap-p1-plot
#| fig-cap: Static square plot with `tmap`
#| out.width: 35%
tmap_mode("plot")
tm_shape(square) +
    tm_borders("black", lwd = 2) 
```


## Style / Theme

As opposed to `ggplot2` where a set of formats is called a "theme", in `tmap`, they are called a "style". You can see the styles available by running:

```{r}
#| label: plotting-tmap-p1-themes
#| message: true
tmap::tmap_style()
```

A style can be set for the duration of the programming session via `tmap_style("<style>")`. Note the difference between `tmap_style()` and `tm_style()`.

```{r}
#| label: fig-plotting-tmap-p1-themes-plot
#| fig-cap: Static square plot with `tmap` using different styles.
see_tmap <- function(style){
    (square |> 
    tm_shape() +
    tm_polygons() +
    tm_style(style)) 
}

theme_1 <- see_tmap(style = "white") 
theme_2 <- see_tmap(style = "classic")
theme_3 <- see_tmap(style = "albatross")
theme_4 <- see_tmap(style = "cobalt")
theme_5 <- see_tmap(style = "watercolor_v3")
theme_6 <- see_tmap(style = "gray")

tmap_arrange(theme_1, theme_2, theme_3, theme_4, theme_5, theme_6, ncol = 3)
```


## Layering

Like `ggplot2`, `tmap` creates plots by layering geoms on the base plot using the `+` operator. As an illustration, a set of simple features will be created to mimic a typical map: a country, a boundary, and three cities. The name of the country is "Square" and it has three cities: "a", "b", and "c".


```{r}
library(ggplot2)
sf::st_polygon(list(rbind(c(0,0), c(0,1), c(1,1),
                          c(1,0),  c(0,0)))) %>%
st_sfc(., crs = 4326) %>%
st_sf(name = "Square", geometry = .) -> country

country %>% st_boundary(.) -> boundary

st_sfc(st_point(c(.5, .1)),
       st_point(c(.8, .9)), 
       st_point(c(.2, .4)), 
       crs = 4326) %>% 
st_sf(name = letters[1:3], .) -> cities

country %>% st_centroid(.) -> center
```

```{r}
#| label: fig-plotting-tmap-layer-plot
#| fig-cap: Creation of static square plot via layering.
tmap_mode("plot")
tmap_style("white")
p1 <- tm_shape(country) + tm_polygons(fill = "white") 
p2 <- tm_shape(country) + tm_polygons(fill = "antiquewhite")
p3 <- tm_shape(country) + tm_polygons(fill = "antiquewhite", lwd = 4) 
p4 <- p3 + tm_shape(cities) + tm_dots(col = "black", size = .5)
p5 <- p4 + tm_shape(cities) + tm_text("name", col = "black", size = 1.25, xmod = -1)
p6 <- p5 + tm_shape(center) + tm_text("name", size = 1.5) 
p7 <- p6 + tm_layout(bg.color = "skyblue")
tmap_arrange(p1, p2, p3, p4, p5, p6, p7, ncol = 3)
```



# Second Plot - Square Land

## Create

Like the `ggplot2` example, this second example will begin to resemble an actual map. First, Africa will be added as a base layer via the `rnaturalearth` package. Then, two `sf` class squares will be created. These squares/islands are around the intersection of the Equator and the Prime Meridian. Then, a mountain, a lake, the Equator and Prime Meridian will be added along with the accompanying text and labels. Finally, the background will be colored blue to resemble water.

```{r}
afrc <- rnaturalearth::ne_countries(continent = "Africa")

# create simple features
land_1 <- st_polygon(list(rbind(c(-10, -10), c(-10, -15), c(-15, -15), c(-15, -10), c(-10, -10))))
land_2 <- land_1 * .75 + 7
lake <- land_1 * .25
mountain <- st_centroid(land_1 - 1)
pm <- st_linestring(matrix(c(0, -20, 0, 20), ncol = 2, byrow = T))
eq <- st_linestring(matrix(c(-20, 0, 20, 0), ncol = 2, byrow = T))

# sfc
land_sfc = st_sfc(list(land_1, land_2))
lake_sfc = st_sfc(lake)
mountain_sfc <- st_sfc(mountain)
pm_sfc <- st_sfc(pm)
eq_sfc <- st_sfc(eq)

# set CRS
land_sfc_wgs = st_sfc(land_sfc, crs = "OGC:CRS84")
lake_sfc_wgs = st_sfc(lake_sfc, crs = "OGC:CRS84")
mountain_sfc_wgs <- st_sfc(mountain_sfc, crs = "OGC:CRS84")
pm_sfc_wgs <- st_sfc(pm_sfc, crs = "OGC:CRS84")
eq_sfc_wgs <- st_sfc(eq_sfc, crs = "OGC:CRS84")

# add attributes
country <- c("A", "B")
land_sfc_wgs_df = st_sf(country = country, land_sfc_wgs)
name <- "Mt. High"
mountain_sfc_wgs_df = st_sf(name = name, mountain_sfc_wgs)
name <- "Lake Clear"
lake_sfc_wgs_df <- st_sf(name = name, lake_sfc_wgs)
name <- "Equator"
eq_sfc_wgs_df <- st_sf(name = name, eq_sfc_wgs)
name <- "Prime Meridian"
pm_sfc_wgs_df <- st_sf(name = name, pm_sfc_wgs)
```


## Plot

From the length of the code below, you can see that creating and plotting maps can be a lengthy and tedious process. Label/text placement can take a lot of time. While there are some improvements to be made, I think the plot encapsulates the process and gives you a strategy to plot your own. You'll notice that this code is not any shorter than the code in the `ggplot2` code.

```{r}
#| label: fig-plotting-tmap-p2-plot
#| fig-cap: Static square land plot with `tmap`
# identify bb in narrative
library(tmap)
library(tmaptools)
tmap_options_reset()
afrc <- rnaturalearth::ne_countries(continent = "Africa")
box <- tmaptools::bb(afrc, xlim = c(-20, 15), ylim = c(-20, 10))
afrc <- afrc[-which(sf::st_is_valid(afrc) == FALSE), ]
squareland <-
    tm_shape(afrc,  bb = box) +
        tm_polygons(fill = "antiquewhite") +
        tm_borders(col = "gray50", lwd = .25) +
    tm_shape(land_sfc_wgs_df) +
        tm_borders("gray50", lwd = .25) +
        tm_polygons("country", fill = "antiquewhite") +
        tm_text("country", ymod = 1) +
    tm_shape(lake_sfc_wgs) +
        tm_polygons(fill = "lightblue") +
        tm_text("Lake Clear", ymod = -1.3, xmod =-1) +
    tm_shape(mountain_sfc_wgs_df) +
        tm_symbols(size = 1, fill = "red", shape = 17) +
        tm_text("Mt. High", ymod = -.75, xmod = 1) +
    tm_shape(eq_sfc_wgs_df) +
        tm_lines(lty = "dotted") +
        tm_text("Equator", xmod = -15, ymod = .5, fontface = "italic", alpha = .5, size = .75) +
    tm_shape(pm_sfc_wgs_df) +
        tm_lines(lty = "dotted") +
        tm_text("Prime Meridian", ymod = -14, xmod = -.5, fontface = "italic",
                alpha = .5, size = .75, angle = 90) +
    tm_xlab("Longitude") +
    tm_ylab("Latitude", rotation = 90) +
    tm_grid(labels.show = T,  alpha = .5, labels.size = .7) +
    tm_compass(position = c(.75, .25)) +
    tm_scalebar(breaks = c(0, 200, 400, 600, 800),
                position = c(.65, .1)) +
    tm_title_in("Square Land",
                size = 1.4,
                padding = c(.5, 1, .5, 1),
                position = tm_pos_in("left", "top"),
                bg.color = "white",
                frame = "black",
                frame.lwd = 2) +
    tm_layout(
              bg.color = "lightblue",
              inner.margins = 0,
              frame = "grey50",
              frame.lwd = 3,
              frame.double.line = F,
              outer.bg.color = "#00000000"
              )
squareland
```




# Key Points

- `tmap` syntax is similar to `ggplot2`

- `tmap` has two modes: "plot" and "view" 

- Use `tmap_style()` to set the style of the plot

- Layers usually begin with a call to `tm_shape()`

<hr>
:::{#refs-plotting-040-rspatial}
:::
:::{#refs-plotting-040-packages}
:::