---
title: "Basics"
fig-height: 3.5
---
```{r}
#| label: plotting-leaflet-intro-libs-load
#| include: false
library(dplyr)
library(magrittr)
library(sf)
library(leaflet)
library(leaflet.providers)
```

# Objectives

The objective for this section is to introduce `leaflet` functions to create an interactive map.

# Introduction

Leaflet is an open-source JavaScript library for interactive maps. It is the behind-the-scenes language for the interactive view mode in `tmap` or it can be accessed directly within `R` using the `leaflet` package. `leaflet` was created by the RStudio team and is available from CRAN. The package documentation is available at [https://rstudio.github.io/leaflet/](https://rstudio.github.io/leaflet/). Here, we will explore the basics of `leaflet` and how to create interactive maps in R. Some of the syntax may be familiar for those familiar with `tmap`. "If you are interested in making maps or in data visualization, Leaflet.js is the library to learn" [@crickard_leaflet_2014].

# Packages Required

```{r}
#| label: plotting-leaflet-intro-libs-show
#| eval: FALSE
library(dplyr)
library(magrittr)
library(sf)
library(leaflet)
library(leaflet.providers)
```

# Call map 

```{r}
#| label: fig-plotting-leaflet-call-map
#| fig-cap: "Leaflet map with default tiles."
leaflet() %>% 
    addTiles()
```


##  Set View

```{r}
#| label: fig-plotting-leaflet-set-view
#| fig-cap: "Leaflet map with Boston, Massachusettes, as center."
leaflet() %>% 
    addTiles() %>% 
    setView(-71.0382679, 42.3489054, zoom = 16)

```

## Add Marker

```{r}
#| label: fig-plotting-leaflet-add-marker
#| fig-cap: "Leaflet map with marker at center"
leaflet() %>% 
    addTiles() %>% 
    setView(-71.0382679, 42.3489054, zoom = 16) %>% 
    addMarkers(-71.0382679, 42.3489054)
```

## Add Popup

```{r}
#| label: fig-plotting-leaflet-add-popup
#| fig-cap: "Leaflet map with marker at center and popup."
leaflet() %>% 
    addTiles() %>% 
    setView(-71.0382679, 42.3489054, zoom = 16) %>% 
    addMarkers(-71.0382679, 42.3489054, popup = "Center of map!")
```


## Add Popup Text

```{r}
#| label: fig-plotting-leaflet-add-popup-text
#| fig-cap: "Leaflet map with marker at center and popup text. Note the use of HTML tags."
leaflet() %>% 
    addTiles() %>% 
    setView(-71.0382679, 42.3489054, zoom = 16) %>% 
    addMarkers(-71.0382679, 42.3489054, 
               popup = paste(
                   "<strong> Value # 1: </strong>", "7", "<br>",
                   "<strong> Value # 2: </strong>", "3", "<br>",
                   "Center of map!"))
```

# Options 

## Remove Zoom Controls

```{r}
#| label: fig-plotting-leaflet-remove-zoom-controls
#| fig-cap: "Leaflet map with zoom controls removed. Note the +/- box is no longer in the left, upper corner of the map."
leaflet(options = leafletOptions(zoomControl = FALSE)) %>% 
    addTiles() %>% 
    setView(-71.0382679, 42.3489054, zoom = 16) %>% 
    addMarkers(-71.0382679, 42.3489054, popup = "Center of map!")
```


## Freeze Zoom

```{r}
#| label: fig-plotting-leaflet-freeze-zoom
#| fig-cap: "Leaflet map with zoom frozen at 16. Note that both the `minZoom` and `maxZoom` are set to 16."
leaflet(options = leafletOptions(zoomControl = FALSE,
                                 minZoom = 16, 
                                 maxZoom = 16)) %>% 
    addTiles() %>% 
    setView(-71.0382679, 42.3489054, zoom = 16) %>% 
    addMarkers(-71.0382679, 42.3489054, popup = "Center of map!")
```

## Dragging

```{r}
#| label: fig-plotting-leaflet-dragging
#| fig-cap: "Leaflet map with dragging disabled. Interactivity is wholly disabled with the exception of the popup."
leaflet(options = leafletOptions(zoomControl = FALSE,
                                 minZoom = 16, 
                                 maxZoom = 16,
                                 dragging = FALSE)) %>% 
    addTiles() %>% 
    setView(-71.0382679, 42.3489054, zoom = 16) %>% 
    addMarkers(-71.0382679, 42.3489054, popup = "Center of map!")
```

## Set Max Bounds

```{r}
#| label: fig-plotting-leaflet-set-max-bounds
#| fig-cap: "Leaflet map with max bounds set. The map will not move outside of the bounds set. The user can zoom in, but is confined to the boundaries set."
leaflet(options = leafletOptions(zoomControl = FALSE,
                                 minZoom = 14, 
                                 maxZoom = 16,
                                 dragging = TRUE)) %>% 
    addTiles() %>% 
    setView(-71.0382679, 42.3489054, zoom = 14) %>% 
    addMarkers(-71.0382679, 42.3489054, popup = "Center of map!") %>%
    setMaxBounds(lng1 = -71.13, lat1 = 42.24, lng2 = -70.93, lat2 = 42.44)
```

# Tiles

Tiled maps are also known as "slippy" maps because they allow for zooming and panning. The map is divided into a grid of tiles and the tiles are loaded as the user zooms in or out. The default tiles in the `leaflet` package are from [OpenStreetMap](https://www.openstreetmap.org/). However, there are many other tile providers available. On the Leaflet website, there is a list of providers on the "Plugins"  [page](https://leafletjs.com/plugins.html#basemap-providers). The `addProviderTiles()` function is used to add tiles to a map. The function requires the name of the provider. The `leaflet.providers` package has a list of providers available and the different tiles can be previewed at [https://leaflet-extras.github.io/leaflet-providers/preview/](https://leaflet-extras.github.io/leaflet-providers/preview/).

## Light

```{r}
#| label: fig-plotting-leaflet-tiles-light
#| fig-cap: "Leaflet map with light tiles."
leaflet(options = leafletOptions(zoomControl = FALSE,
                                 minZoom = 14, 
                                 maxZoom = 16,
                                 dragging = TRUE)) %>% 
    addProviderTiles("Esri.WorldGrayCanvas") %>% 
    setView(-71.0382679, 42.3489054, zoom = 14) %>% 
    addMarkers(-71.0382679, 42.3489054, popup = "Center of map!") %>%
    setMaxBounds(lng1 = -71.13, lat1 = 42.24, lng2 = -70.93, lat2 = 42.44)

```

## Dark

```{r}
#| label: fig-plotting-leaflet-tiles-dark
#| fig-cap: "Leaflet map with dark tiles."
leaflet(options = leafletOptions(zoomControl = FALSE,
                                 minZoom = 14, 
                                 maxZoom = 16,
                                 dragging = TRUE)) %>% 
    addProviderTiles("Stadia.AlidadeSmoothDark") %>% 
    setView(-71.0382679, 42.3489054, zoom = 14) %>% 
    addMarkers(-71.0382679, 42.3489054, popup = "Center of map!") %>%
    setMaxBounds(lng1 = -71.13, lat1 = 42.24, lng2 = -70.93, lat2 = 42.44)

```


## WMS

Another type of tile layer that can be added to a map is called Web Mapping Service (WMS) tile layer. WMS tiles are requested and transferred images over the web using HTTP [@crickard_leaflet_2014]. The function in `R` is `addWMSTiles()` and it requires the URL of the WMS server and the layer name. The `leaflet` documentation states that, with regard to the layers name, they are defined in the "GetCapabilities" XML document; however, it can be difficult to understand.  `leaflet` recommends using software like QGIS to see what layers are available.

```{r}
#| label: fig-plotting-leaflet-tiles-wms
#| fig-cap: "Leaflet map with WMS tiles. These tiles are from the United States Geographic Service."
leaflet() %>% 
     setView(-71.0382679, 42.3489054, zoom = 14) %>% 
    addWMSTiles(
        "https://basemap.nationalmap.gov/arcgis/services/USGSTopo/MapServer/WmsServer",
        layers = "0",
        options = WMSTileOptions(format = "image/png", transparent = TRUE),
        attribution = "USGS"
    ) 
```


## Options

```{r}
#| label: fig-plotting-leaflet-tiles-options
#| fig-cap: "Leaflet map with tile options set."
leaflet() %>% 
    addProviderTiles("Esri.WorldGrayCanvas",
                     options = providerTileOptions(minZoom = 14, maxZoom = 16, attribution = "source")
    ) %>%
    setView(-71.0382679, 42.3489054, zoom = 14) %>% 
    addMarkers(-71.0382679, 42.3489054, popup = "Center of map!") %>%
    setMaxBounds(lng1 = -71.13, lat1 = 42.24, lng2 = -70.93, lat2 = 42.44)
```

# Example

## Create

```{r}
#| label: fig-plotting-leaflet-example-create
sf::st_polygon(list(rbind(c(-71.13, 42.24), c(-71.13, 42.44), c(-70.93, 42.44), c(-70.93, 42.24), c(-71.13, 42.24)))) %>% 
sf::st_sfc() %>% 
sf::st_sf(name = "district") -> poly

lng <- seq(from = -71.13, to = -70.93, by = .01)
lat <- seq(from = 42.24, to = 42.44, by = .01)
points <- data.frame(lng = sample(lng, size = 10),
                     lat = sample(lat, size = 10),
                     name = letters[1:10])
```

## Plot

```{r}
#| label: fig-plotting-leaflet-example-plot
#| fig-cap: "Leaflet map with polygons, markers, and circles."
leaflet() %>% 
    addProviderTiles("Esri.WorldGrayCanvas",
                     options = providerTileOptions(
                         minZoom = 10, 
                         maxZoom = 16, 
                         attribution = "source")
                     ) %>%
    setView(-71.0382679, 42.3489054, zoom = 10) %>% 
    addMarkers(-71.0382679, 42.3489054, popup = "Center of map!") %>%
    setMaxBounds(lng1 = -71.13, lat1 = 42.24, lng2 = -70.93, lat2 = 42.44) %>%
    addPolygons(data = poly, popup = ~name, color = "red", weight = 2, opacity = 1, fill = TRUE, fillColor = "blue", fillOpacity = 0.2) %>%
    addCircleMarkers(data = points, lng = ~lng, lat = ~lat, popup = ~name, radius = 10, color = "green", fillColor = "yellow", fillOpacity = 0.5)

```

# Layer Groups

```{r}
#| label: fig-plotting-leaflet-example-layer-groups
#| fig-cap: "Leaflet map with layer control."
leaflet() %>% 
    # base groups
    addTiles(group = "Default") %>%
    addProviderTiles("Esri.WorldGrayCanvas", group = "Light") %>% 
    addProviderTiles("Stadia.AlidadeSmoothDark", group = "Dark") %>%
    addWMSTiles(
        "https://basemap.nationalmap.gov/arcgis/services/USGSTopo/MapServer/WmsServer",
        layers = "0",
        options = WMSTileOptions(format = "image/png", transparent = TRUE),
        attribution = "USGS",
        group = "WMS"
    ) %>% 
    # view/bounds
    setView(-71.0382679, 42.3489054, zoom = 10) %>%
    setMaxBounds(lng1 = -71.13, lat1 = 42.24, lng2 = -70.93, lat2 = 42.44) %>%
    # add features
    addMarkers(-71.0382679, 42.3489054, 
               popup = "Center of map!",
               group = "Marker") %>%
    addPolygons(data = poly, popup = ~name, color = "red", weight = 2, 
                opacity = 1, fill = TRUE, fillColor = "blue", 
                fillOpacity = 0.2,
                group = "Polygon") %>%
    addCircleMarkers(data = points, lng = ~lng, lat = ~lat, 
                     radius = 10, color = "green", 
                     fillColor = "yellow", fillOpacity = 0.5,
                     popup = ~paste0(
                         "<b> Letter: </b>", name, "<br>"
                     ),
                     group = "Circles") %>% 
    # layer control
    addLayersControl(baseGroups = c("Default", "Light", "Dark", "WMS"),
                     overlayGroups = c("Marker", "Polygon", "Circles"),
                     options = layersControlOptions(collapsed = FALSE))
```

# Key Points

- `leaflet` is an open-source JavaScript library for interactive maps.

- The `setView()` function is used to set the center of the map and the zoom level.

- The zoom and pan features can be turned off with the `leafletOptions()` function.

- While the Open Street Map tiles are the default, there are many tiles available to customize a map.

- Another way to introduce interactivity is through layer groups.  It can allow the user to decide what they want to see on the map.

<hr>
:::{#refs-plotting-051-rspatial}
:::
:::{#refs-plotting-051-packages}
:::





